---
title: "Spam_Detection_Classification"
author: "Jess Chen"
format: gfm
editor: visual
---

## Project 4 - Spam Detection Model

In this project, we build a spam detection model using native Bayes model. The data used to train and test the model was sourced from an open sourced project called, Apache SpamAssassin (link in references). The process included uploading the raw data into github, uploading and cleaning the data in R, categorizing the data before training the model and ultimately, testing and refining the model.

### 1. Setup & Upload

We firstly setup libraries that will be used for this project including tidytext, caret and dplyer. Followed by uploading the documents into github.

```{r}
library(tm)        # Text mining
library(caret)     # ML framework
library(e1071)     # Naive Bayes
library(tidytext)
library(dplyr)
library(httr)


base_url <- "https://raw.githubusercontent.com/JessChen0/spam-detection-program/refs/heads/main/"

# Function to read all emails from a GitHub folder
read_emails_from_github <- function(folder_name) {
  # Get the list of files using the GitHub API
  api_url <- paste0("https://api.github.com/repos/JessChen0/spam-detection-program/contents/", folder_name)
  
  response <- GET(api_url)
  files <- content(response)
  
  # Download and read each file
  emails <- c()
  for (file in files) {
    raw_url <- file$download_url
    text <- paste(readLines(url(raw_url), warn = FALSE), collapse = " ")
    emails <- c(emails, text)
  }
  return(emails)
}

# Load ham and spam emails
cat("Loading ham emails... (this may take a minute)\n")
ham_emails  <- read_emails_from_github("easy_ham")

cat("Loading spam emails... (this may take a minute)\n")
spam_emails <- read_emails_from_github("spam")

cat("Done! Ham:", length(ham_emails), "emails | Spam:", length(spam_emails), "emails\n")
```

Check emails are uploaded.

```{r}
# Check how many emails loaded
cat("Ham emails:", length(ham_emails), "\n")
cat("Spam emails:", length(spam_emails), "\n")

# Preview the first ham email
ham_emails[1]

# Preview the first spam email
spam_emails[1]


```

### 2. Data Cleanup and Label

Next, we will combine the emails and label them

```{r}

# Label each email
ham_labels  <- rep("ham", length(ham_emails))
spam_labels <- rep("spam", length(spam_emails))

# Combine into one data frame
emails_df <- data.frame(
  text  = c(ham_emails, spam_emails),
  label = c(ham_labels, spam_labels),
  stringsAsFactors = FALSE
)

# Convert label to a factor 
emails_df$label <- as.factor(emails_df$label)

# Function to remove dodgy characters before processing
clean_text <- function(x) {
  x <- iconv(x, from = "UTF-8", to = "ASCII", sub = "")  # remove non-ASCII characters
  return(x)
}

# Create corpus
corpus <- VCorpus(VectorSource(emails_df$text))

# Clean the text - note clean_text is added at the top
corpus <- tm_map(corpus, content_transformer(clean_text))  # remove special characters FIRST
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removeWords, stopwords("en"))
corpus <- tm_map(corpus, stripWhitespace)

# Preview
cat("BEFORE cleaning:\n")
print(emails_df$text[1])

cat("\nAFTER cleaning:\n")
print(corpus[[1]]$content)



```

### 3. Covert Emails and Count Frequency

Count and capture word frequency profile for each email. This will help inform the classifier later on how to categorize the "new" emails coming in.

```{r}
# Create the Document-Term Matrix
dtm <- DocumentTermMatrix(corpus)

# Check its size
cat("DTM dimensions:\n")
dim(dtm)

# Remove words that appear in less than 1% of emails
dtm <- removeSparseTerms(dtm, 0.99)

# Check the new size
cat("DTM dimensions after trimming:\n")
dim(dtm)



```

### 4. Train Model

We will now split the dataset into training ds and testing ds. But first, we have to shuffle the data and will set a seed to do this.

```{r}
# Convert word counts to yes/no
train_data_bool <- train_data
test_data_bool  <- test_data

train_data_bool[, -ncol(train_data)] <- train_data[, -ncol(train_data)] > 0
test_data_bool[, -ncol(test_data)]   <- test_data[, -ncol(test_data)] > 0

# Retrain the model
cat("Retraining model...\n")
model_bool <- naiveBayes(label ~ ., data = train_data_bool)
cat("Done!\n")

# Test it
predictions_bool <- predict(model_bool, test_data_bool)
confusionMatrix(predictions_bool, test_data_bool$label, positive = "spam")

```

Finally, we will train the Naive Bayes Classifier!

```{r}
# Train the Naive Bayes model
cat("Training the model... (this may take a minute)\n")
model <- naiveBayes(label ~ ., data = train_data)
cat("Done!\n")


```

### 5. Test Model

With the model built, we can now test to see how well it does!

```{r}

# Use the model to predict labels on the test set
predictions <- predict(model, test_data)

# Compare predictions to the real labels
confusionMatrix(predictions, test_data$label, positive = "spam")

```

### 6. Refine Model

The confusion matrix above highlighted the dangers of false positives. In this situation, it is detrimental to falsely identify a ham email as spam. Even though we correctly identified all spam, we also allowed a lot of real emails to potentially not be reviewed!

In this case, we iterated the model to increase the confidence level to 90% for spams. This allowed us to correctly identify spam but also, the model is being more precise and leaning on the side of more emails be labelled ham, which as it turns out leads to more correct answers.

```{r}
# Convert to factor
train_data_factor <- as.data.frame(lapply(train_data_bool[, -ncol(train_data_bool)], factor))
test_data_factor  <- as.data.frame(lapply(test_data_bool[, -ncol(test_data_bool)], factor))
train_data_factor$label <- train_data_bool$label
test_data_factor$label  <- test_data_bool$label

# Train model
cat("Training model...\n")
model_factor <- naiveBayes(label ~ ., data = train_data_factor)
cat("Done!\n")

# Test with strict threshold
probabilities      <- predict(model_factor, test_data_factor, type = "raw")
threshold          <- 0.90
predictions_strict <- ifelse(probabilities[, "spam"] > threshold, "spam", "ham")
predictions_strict <- factor(predictions_strict, levels = c("ham", "spam"))
confusionMatrix(predictions_strict, test_data_factor$label, positive = "spam")

```

### 7. Conclusion

In this assignment, we leveraged Claude LLM to help build out a classification model to identify spam emails. We used native Bayes to train the model on 1200 emails and tested on 300. Our initial model resulted in far too many false positives (e.g., where a ham is incorrectly labelled spam), this would be harmful in the case of a spam filter if the reader then ignores important emails. To rectify this, we retooled the model to use a stricter confidence level to filter out less spam, therefore allowing for more ham emails to be correctly labelled.

### 0. References

Raw model training data: <https://spamassassin.apache.org/old/publiccorpus/>

Claude Sonnet 4.6 (2026) Anthropic LLM. Link to chat: <https://claude.ai/share/4308269d-a184-4199-9355-41b758434fc0>
